PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-05-16 IEEEtran/paper.tex
Missing sigdigits result in Table 1 for L1 Penalty
Edits and comments

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model , allowing for tool evaluation through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automated, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\addCite{\cite{higham2002accuracy}}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\addCite{\cite{goubault2006static}} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses (a) by relying on
Fuzzy~\addCite{\cite{kiar2020comparing}}, a pre-existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
classes through meta-programming, (3) an interactive Plotly dashboard to
visualize stability measures throughout program executions. }

This manuscript presents the design of \pytracer and demonstrates its potential
in various use cases. Through stability evaluations of the SciPy and
scikit-learn libraries, we demonstrate \pytracer as a practical solution to
review the numerical stability of extensive code bases.

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \remove{preservers} \add{preserves} the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with \remove{decorator-based approaches} \add{a decorator-based
    approach~\mbox{\cite{hunt2019decorators}}}. In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to instrument, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they,} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. PyTracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis [23], and random
seed analysis [11].  \add{PyTracer is agnostic to the noise model used and can therefore
be used for all these methods}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{These libraries are written in large parts in a 
compiled language other than Python (such as C or Cython) and their execution is not 
perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
to provide an overall picture of numerical uncertainty in the tested examples. 
Depending on the use-case of interest, one might choose to use MCA only in 
a subset of these libraries or even just in the Python interpreter.} \add{We
    used MCA in Random Rounding (RR) mode}. We used the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on
SciPy and scikit-learn examples. Columns \textit{trace} and \textit{parse}
report on the tracing and aggregation steps with two possible outcomes:
\add{green \valid\xspace for success and red \cross\xspace for error. Tracing errors 
occur in the application as a result of 
numerical perturbations, whereas parsing errors occur in \pytracer when traces result from different 
control flow paths. 
} \remove{, and 3)
    orange for errors due to the Fuzzy noise injection that are
    not considered numerical errors, meaning that the execu-
    tion raised a Python error but due to a perturbation that
    should not have been introduced (i.e. a floating-point value
    representing an integer).}
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples\remove{ and showed an average
    precision loss of 8 bits}. The remaining examples failed due to numerical errors
in the libraries, which we discuss hereafter. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the \add{result} precision for the main outputs of SciPy and scikit-learn examples. We
measured the \add{result} precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produces many outputs, the average value among them. \add{Significant bits are computed on the final program output 
and can therefore be computed even when parsing fails.} \add{We use the term ``result precision"
to denote a statistical variability measured in number of significant bits, in 
contrast with the other meanings of the word precision used in expressions
``virtual precision" or ``single/double precision'' in this paper.} The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

\noindent \remove{\textbf{3.2 Effect of MCA modes}}

\remove{
    Random Rounding perturbs only the result of a floating-point operation while
    Full MCA perturbs both the inputs and the result. Therefore, Full MCA may
    trigger catastrophic cancellations while Random Rounding only introduces
    round-off errors which generally have a lesser impact. Moreover, Random Rounding
    preserves exact operations --- operations where the result can be exactly
    represented at the virtual precision --- while Full MCA does not.}

\remove{\pytracer can only aggregate traces obtained from the same control flow path. In
    particular, program branches triggered by an unstable floating-point comparison
    lead to different control flows and result in parsing errors. For example,
    comparing a residual error or the distance between two points to a strict
    threshold in an iterative scheme generally results in unstable branches. Full
    MCA even perturbs exact values such as integers represented with floating-point
    values, leading to many execution failures. For instance, the \texttt{fft1}
    example from SciPy raised an \texttt{ValueError: operands could not be broadcast
        together with shapes (601,) (600,)} error due to an array shape mismatch. A
    closer inspection showed that one of the arrays involved in the error was
    created with the NumPy function \texttt{linspace} that itself calls function
    \texttt{arange} (Listing~\ref{fig:pyarray_range_code}). Lines 3174-3178 show
    that the function stores the array size in a floating-point value that will be
    perturbed by Full MCA, as shown in Table~\ref{tab:mca_result_linspace},
    resulting in array sizes that fluctuate between N and N+1. It is worth noting
    that 78\% of execution issues encountered in Full MCA mode instrumentation are
    due to this type of ``typing" errors.  
    Those code regions should not be perturbed by MCA since the injected noise does
    not make sense.}

@@

\add{PyTracer supports native floating point formats as well as compound data types,
    such as complex floating point values, N-dimensional arrays and arbitrary Python
    classes.}
Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs} \add{show the
    visualization of the \texttt{fft\_1D} input and output values. Results are
    complex values encoded in double precision. PyTracer allows visualizing the
    \add{result} precision of real (blue points) and imaginary (orange points) parts.}
\add{The} top row in Figures~\ref{fig:fft1D_inputs} and~\ref{fig:fft1D_outputs}
show the input mean value while the bottom row shows the standard deviation. \add{Even though we 
used Random Rounding (RR), function inputs are perturbed since they result from multiple intermediate 
computations perturbed with RR.} The
points with low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal noise
introduced by MCA on the inputs is in the order of $10^{-14}$, two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

@@

\add{We illustrate on two examples how PyTracer is effective for analyzing
    numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
\add{When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.}
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we see that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
non-significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    According to Equation}~\ref{eq:sig-digits}, \add{a number of significant bits less than 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

\add{Similarly, for OLS, PyTracer tracers show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
    the \texttt{NaN} value disapears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a \add{result} precision of
    42 bits in average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues,
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
    \add{result} precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

% \begin{figure}
%     \centering
%     \includegraphics[width=\linewidth]{figure/parsing_time.pdf}
%     \caption{\pytracer time to aggregate traces.}
%     \label{fig:performance_parsing_time}
% \end{figure}

% \begin{figure}
%     \centering
%     \includegraphics[width=\linewidth]{figure/parsing.pdf}
%     \caption{\pytracer trace and aggregation files size.
%     The figure presents the size measured for the trace and the aggregation size files.}
%     \label{fig:performance_parsing_size}
% \end{figure}

@@

% \begin{table}
% \centering
% \begin{subfigure}[t]{.75\linewidth}
%     \centering
%     \begin{tabular}{|l|c|c|c|c|}
%     \hline 
%     Application & Original & Pytracer only & RR & MCA \\
%     \hline 
%     Loading & 0.456 & 8.152 & - & - \\
%     \hline
%     discrete\_cosine\_transform & 0.280 & 8.176 & 10.203 & - \\ 
%     fft\_1D & 0.650 & 9.075 & 11.345 & - \\
%     fft\_2D & 0.311 & 9.500 & 11.929  & 16.433   \\
%     \hline
%     interpolation\_1D & 0.377 & 8.803  & 11.954  & -   \\
%     multivariate\_data &0.626 & 8.648 &  12.711  & - \\
%     bspline & 0.109  & 9.050 & 13.396  & 16.830   \\
%     spline\_1D & 0.111 & 8.618 & 10.960   & - \\
%     spline\_2D & 0.655 & 8.649 & 19.424  & - \\
%     \hline
%     broyden\_fletcher\_goldfarb\_shanno &0.475 & 8.292 & 11.033 & 13.685 \\
%     global\_optimization &1.257 & 55.916  & 93.322  & - \\
%     slsqp & 0.469 & 8.192 & 10.795 & 12.439  \\
%     least\_square\_minimization & 0.505 & 10.151  & 13.888 & -  \\
%     nelder\_mead\_simplex & 0.471 & 8.418  & 11.382  & 12.792  \\
%     newton\_conjugate\_gradient & 0.487 & 8.548  & 11.956  & 14.281 \\
%     root\_findings & 0.462 & 8.198  & 10.787   & - \\
%     root\_finding\_large & 0.801 & 11.525 & 41.493   &  82.701  \\
%     root\_finding\_large\_preconditioned & 0.702 & 10.580 & 33.199   & 57.579  \\
%     trust\_region\_exact & 0.482 & 8.170 & 10.400  & 12.524 \\
%     trust\_region\_truncated\_generalized\_lanczos & 0.459 & 8.457 & 11.216 & 13.245  \\
%     trust\_region\_newton\_conjugate\_gradient & 0.479 & 8.538 & 11.179 & 13.488 \\
%     \hline
%     Adaboost &1.388 &14.553 &  20.999 & - \\
%     Bayesian Ridge Regression & 1.086 & 10.344 & 90.299 & -  \\
%     Online classifier comparison & 15.964 & 72.691 & 2104.114 & -  \\
%     K-Means Clustering & 1.621 & 17.614  & 34.005 & 52.902   \\
%     Covariance estimation & 1.357 & 11.755 & 23.964  & -   \\
%     Decision Tree Regression & 1.185 & 11.194 & 15.760 & 15.918  \\
%     Digits Classification & 1.672 & 12.725 & 40.634 & - \\
%     Face Recognition & 32.159  & 74.356 & - & - \\
%     L1 Penalty & 1.568 & 12.591 & 65.547 & 128.220 \\
%     Lasso and elastic net & 1.057 & 9.903 & 17.078 & -  \\
%     Separating hyperplan & 1.376 & 11.532  & 22.333 & 25.036 \\
%     MNIST classification & 19.650 & 37.838 & 402.389  & - \\
%     Multitask Lasso & 1.090 &10.196 & 15.025  & - \\
%     Orthogonal matching & 1.161 & 12.129 & 21.779   & 31.135 \\
%     PCA decomposition & 1.239 & 10.642 & 15.520 & 17.952    \\
%     Robust Linear Regression & 1.370 & 11.144  & 14.280  & - \\
%     Segmentation toy &1.679 & 22.385 & 87.811  & - \\
%     SVM & 1.403 & 11.737 & 23.203 & 24.126   \\
%     Tomography & 11.108 & 40.177  & 1376.134 & -  \\
%     Weighted samples & 1.358 & 11.768 & 21.255 & 20.541   \\
%     \hline
%     \end{tabular}
% \end{subfigure}
%     \caption{\pytracer tracing overhead in seconds. \tristan{what is the unit in this table?}}
%     \label{tab:pytracer_overhead}
% \end{table}

% \begin{table}[]
%     \centering
%     \begin{tabular}{c|c|c|c|c}
%          Pytracer & Verificarlo Instrumentation & Backend & Time (hours:minutes:seconds) & Overhead \\
%          \hline 
%          No & No & - & 00:30:51 & 1 \\
%          No & Yes & IEEE & 01:32:56 & 3 \\
%          No & Yes & RR & 16:52:08 & 32 \\
%          Yes & No & - & 03:11:07 & 6  \\
%          Yes & Yes & IEEE & 11:14:33 & 22 \\
%          Yes & Yes & RR & 25:05:54 & 48 \\
%     \end{tabular}
%     \caption{Overheads of \pytracer on PyAFQ experiment with and without fuzzy.}
%     \label{tab:pytracer_overhead}
% \end{table}

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model, allowing for \add{numerical profiling} through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automated, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\cite{higham2002accuracy}, which require
detailed modeling of the program and its numerical implementation. Similarly,
static code analysis techniques~\cite{goubault2006static} hardly scale
to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses (a) by relying on
Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
classes through meta-programming, (3) an interactive Plotly dashboard to
visualize stability measures throughout program executions. }

This manuscript presents the design of \pytracer and \add{highlights} its potential
in various use cases. Through stability evaluations of the SciPy and
scikit-learn libraries, we demonstrate \pytracer as a practical solution to
review the numerical stability of extensive code bases.

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \add{preserves} the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with \remove{decorator-based approaches} \add{a decorator-based
    approach~\mbox{\cite{hunt2019decorators}}}. In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to instrument, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact in practice since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. Therefore, \pytracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis \TG{[23] fix ref}, and random
seed analysis \TG{fix ref [11]}.  \add{PyTracer is agnostic to the noise model used and can therefore
be used for all these methods}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{These libraries are written in large parts in a 
compiled language other than Python (such as C or Cython) and their execution is not 
perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
to provide an overall picture of numerical uncertainty in the tested examples. 
Depending on the use-case of interest, one might choose to use MCA only in 
a subset of these libraries or even just in the Python interpreter.} \add{We
    used MCA in Random Rounding (RR) mode, with} the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on
SciPy and scikit-learn examples. Columns \textit{trace} and \textit{parse}
report on the tracing and aggregation steps with two possible outcomes:
\add{green \valid\xspace for success and red \cross\xspace for error. Tracing errors 
occur in the application as a result of 
numerical perturbations, whereas parsing errors occur when traces result from different 
control flow paths.} 
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples\remove{ and showed an average
    precision loss of 8 bits}. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the \add{result} precision for the main outputs of SciPy and scikit-learn examples. We
measured the \add{result} precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produce\add{d multiple} outputs, the average value among them. \add{Significant bits were computed on the final program output 
and could therefore be computed even when parsing failed.} \add{We use the term ``result precision"
to denote the statistical variability measured in number of significant bits, in 
contrast with the other meanings of the word precision used in expressions
``virtual precision" or ``single/double precision'' in this paper.} The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\add{PyTracer supports native floating point formats as well as compound
    data types, such as complex floating point values, N-dimensional arrays
    and arbitrary Python classes.} Figures~\ref{fig:fft1D_inputs}
    and~\ref{fig:fft1D_outputs} \add{show the visualization of the
    \texttt{fft\_1D} input and output values. Results are complex values
    encoded in double precision. PyTracer allows visualizing the
    \add{result} precision of real (blue points) and imaginary (orange
    points) parts.} \add{The} top row in Figures~\ref{fig:fft1D_inputs}
    and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
    row shows the standard deviation. \add{Even though Random Rounding
    (RR) was used, function inputs were perturbed since they result from
    multiple intermediate computations perturbed with RR.} The points with
    low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
    near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
    exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
    noise introduced by MCA on the inputs is in the order of $10^{-14}$,
    two orders of
magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 corresponds to the actual frequencies of the input
signal.

@@

\add{We illustrate on two examples how PyTracer is effective for analyzing
    numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
\add{When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.}
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we observed that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.

@@

The second example compares a Bayesian Ridge Regression (BRR) to the Ordinary
Least Squares (OLS) estimator on a synthetic dataset and for one-dimensional
regression using polynomial feature expansion. Although the example does not
raise a runtime error, the regression coefficients computed from the fitting are
not significant. \pytracer traces reveal that the SVD solver used is once again the root
cause for this error. \add{Indeed, the BayesianRidge model first computes an SVD
    before iteratively updating the coefficients until reaching the convergence
    criterion. However, PyTracer shows an issue in the singular values computed.}
\add{We can see in Figure}~\ref{fig:BRR_svd} \add{that the last singular values is
    \texttt{NaN} due to numerical errors in the SVD, although the remaining
    singular values exhibit high \add{result} precision (between 44 and 50 bits).
    Surprinsingly, this \texttt{NaN} value does not raise an exception and is silently
    propagated. Indeed, the stopping criterion is never verified and the 
    optimization stops after reaching the maximal number of iterations.
    Figure}~\ref{fig:BRR_svd_fixed} \add{shows that uses
    the \texttt{gesvd} instead of the \texttt{gesdd} method removes the \texttt{NaN} value
    although the last singular values stills have a low \add{result} precision (below 0).
    According to Equation}~\ref{eq:sig-digits}, \add{a number of significant bits less than 0 means that the
    variability ($\sigma$) is higher than the average value ($\mu$)}.

\add{Similarly, for OLS, PyTracer traces show that the
    regression coefficients fitted are \texttt{NaN} values. By looking at the callgraph, we can
    see that the scipy calls the \texttt{lstsq} method from the
    \texttt{scipy.linalg} package which uses the \texttt{gesdd} LAPACK function.
    Figure}~\ref{fig:OLS_svd} \add{shows that one singular value is a \texttt{NaN} value,
    which corrupted the calculation of the coefficients. When using the \texttt{gesvd} method, 
    the \texttt{NaN} value disapears 
    (Fig.}~\ref{fig:OLS_svd_fixed}\add{), resulting in a \add{result} precision of
    42 bits in average in the coefficients. We can see that having
    access to all the values computed by the application with PyTracer allows
    the user to efficiently identify numerical issues.}

In conclusion, \pytracer and the MCA model were helpful to detect the
instability of the \texttt{gesdd} method mentioned in several GitHub issues \TG{(such as ...)},
including the LAPACK repository. Furthermore, it is worth mentioning that a Pull
Request has recently been submitted \TG{in the ... repo} to let the user select the SVD method and
avoid this kind of issue. Thus, \pytracer can detect instabilities that affect
fundamental scientific computing components such as the LAPACK library. \add{The
    unified view of all traces allows the user to quickly navigate through the
    \add{result} precision of different variables.}
% The Pytracer findings can therefore be extended to many application domains. %
% \tristan{conclusion needs to be discussed. Note to self: generaliser un peu la
% conclusion: pytracer permet de se rendre compte de problemes de stabilite dont
% plusieurs librairies sont dependantes.}

@@

@@
